Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-03-20 doc/invt.tex
update doc

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques in free energy simulations.
  %
  They allow one to sample a flat distribution
  along a quantity of interest, $z$,
  by adaptively constructing a bias potential
  that offsets the potential of mean force (PMF).
  %
  The two algorithms differ by
  the manner of updating the bias potential:
  %
  in the WL case, each simulation step triggers
  an update to only
  the bin containing the current $z$,
  while in the metadynamics case,
  several neighboring bins
  are affected as well.
  %
  The asymptotic error of the resulting PMF
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  a simple and effective formula for the magnitude
  is given by the inverse of the number of simulation steps.
  %
  In this study, we shall reexamine the optimality of this prescription,
  and attempt to give similar prescriptions for the metadynamics case.
  %
  Our results show that the optimal schedule
  and the effectiveness of metadynamics
  are sensitive to the simulation length and other parameters.
  %Further, we compared the two algorithms, and found that
  %although metadynamics produces a smoother PMF profile,
  %it is less effective in reducing errors with the shorter wavelengths.
  %%
  %Thus, the latter relies more heavily on the assumption of
  %the smoothness of the PMF profile,
  %especially with a wide updating window.
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$, and to see how
the distribution changes with external conditions,
such as temperatures and pressures.
%
The negative logarithm $-\log p^*(z)$,
or the potential of mean force (PMF),
defines a free energy.
%
A straightforward approach of computing the PMF
is to run multiple independent simulations
under different conditions,
and then piece together the information
gathered there.
%
This solution is often unsatisfactory
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

A better solution is to artificially alter
the target distribution,
such that we can, in a single simulation, sample
a flat distribution\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}
along $z$.
%
This alleviates the above problem of local trapping
as the system is now forced to travel more flexibly
along the $z$ direction,
which often represents
a direction of slow motion.
%
It, however, requires one to efficiently construct a
bias potential for an often unknown system.
%
Note that the bias potential must exactly cancel the PMF
to achieve a flat distribution.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{huber1994, laio2002, laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is actively built-up on-the-fly
during simulation.
%
In essence, these techniques
encourage the system to leave a region that has been visited
by elevating the bias potential there.
%
A main difference of the two techniques
is that in the WL case, an update is limited to the bin
containing the current $z$,
while metadynamics adopts an extended updating window
covering several neighboring bins.
%
The latter ensures a smooth profile
suitable for numerical differentiation for force calculation,
and hence is more often used molecular dynamics simulations.
%
To be effective, the updates to bias potential
are made at every few steps,
while a side effect is that the underlying dynamics
is no longer a equilibrium sampling
required by statistical mechanics.
%
Thus, one often has to reduce the magnitude of updating
over the course of simulation.

Naturally, the updating schedule,
or the manner of reducing
the updating magnitude\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}
determines the precision of the final bias potential,
hence that of PMF.
%
For the WL algorithm, it is found
that asymptotically the optimal magnitude of updating $\ln f$
should be given by the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008},
where ``time'' is defined as
the number of simulation steps
divided by the number of bins.
%
For metadynamics, ...

In this study,
we shall reexamine the inverse-time formula,
and similarly optimize schedules
for updating schemes involving several neighboring bins,
such as those used in metadynamics.
%
It turns out that the optimal updating schedule
can be sensitive the simulation length.
%
We also shall show that
the single-bin updating scheme used by the WL algorithm
is best in asymptotic convergence.
%
For finite-length simulations, however,
metadynamics may deliver better performance.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

In this section,
we first review the basics and fix notations
in Sec. \ref{sec:background}.
%
We then prove the optimality
of the inverse-time formula
for the single-bin updating WL algorithm
in Sec. \ref{sec:single-bin}.
%
Next, we discuss the generalized
multiple-bin schemes
in Secs. \ref{sec:multiple-bin}
and \ref{sec:band-matrix}.
%

@@

\subsubsection{Flat-distribution sampling}

@@

For a large system,
the distribution $p^*_i$ tends to
be localized around some $i_0$,
%
and to find out the global property,
it is often desirable to carry out
a biased sampling that targets
a wider distribution $p_i$.
%
%Here, we refer to simulations that target
%a flat or nearly flat distribution
%as entropic or multicanonical sampling.

@@

Generally, we may wish to sample
a nearly flat distribution\cite{dayal2004, trebst2004, singh2011},
$p_i$,
by adjusting the bias potential $V_i$
on the fly.
%
For example,
we can monitor the histogram accumulated over
a period of time and use it for $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1},
and then update
$V_i$ according to
$$
V_i^{\mathrm{new}}
=
V_i^{\mathrm{old}}
-
\log \frac{ p_i } { \pi_i }.
$$
%
%From Eq. \eqref{eq:pi_p_phi1},
%it is clear that to achieve the target distribution
%$\pi_i = p_i$ requires
%%
%\begin{equation}
%\log p^*_i
%=V_i + \ln p_i + c
%\label{eq:bias_pmf}
%\end{equation}
%%
%%
%In the end of the simulation,
%the PMF, $-\log p^*_i$,
%can be deduced from Eq. \eqref{eq:bias_pmf}.
%
The above inversion process,
adopted by early multicanonical or
entropic sampling\cite{berg1992, lee1993},
can be inconvenient in practice.
%
For example, a reasonably precise the histogram
requires an accumulation period
longer than the autocorrelation time,
whereas the latter is difficult to
estimate beforehand for an unknown system.

@@

In the WL algorithm\cite{wang2001, wang2001pre},
$v_i$ are updated
in each MC step $t$,
and $v_i(t)$ as a function of $t$
is updated as
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }.
  \label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
We refer to the above scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$,
%
In a more general \emph{multiple-bin scheme}, however,
several neighboring bins are updated:
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i },
  \label{eq:mbin_update}
\end{equation}
%
where $w_{ij}$ are a set of coefficients satisfying
$\sum_{j=1}^n w_{ij} = 1$.
%
We shall discuss other restrictions on $w_{ij}$
in Sec. \ref{sec:updating-matrix}.
%
The multiple-bin scheme is more commonly
used in metadynamics.

@@

In the original WL scheme,
the updating magnitude $\alpha$ (or $\ln f$
in the original paper) is kept as a constant
for a period of time,
which is referred to as a stage below.
%
The histogram collected in the stage is monitored.
%
Once the histogram is sufficiently flat,
we are allowed to enter a new stage
with a reduced $\alpha$\cite{wang2001, wang2001pre}
(usually half as large as
that in the previous stage).
%
This scheme works well for early stages.
%
However, in later stages, it tends to reduce $\alpha$
too quickly, making the asymptotic error
saturate\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008}.
%
Similarly, for metadynamics,
there are various ways of reducing the updating magnitude,
particularly in the development of
self-healing umbrella sampling
and well-tempered metadynamics\cite{
  marsili2006, barducci2008, dickson2011, dama2014}.

@@

A more effective way
of updating the $\alpha(t)$
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$ is the number of steps
from the beginning of the MC simulation,
which shall be referred to as the ``time'' below.
%
This surprisingly simple formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
There are, however, some reservation about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $C/t$ with a constant $C$
different from $1.0$\cite{
morozov2007, zhou2008, morozov2009}.

@@

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(t) = x(0) \, e^{-\lambda \, q(t)}
+ \int_0^t \dot u\bigl( q(t') \bigr) \, \xi(t') \, dt',
\label{eq:xt_solution}
\end{equation}
%
where,
%
\begin{equation}
q(t) \equiv \int_0^t \alpha(t') \, dt',
\label{eq:qt_definition}
\end{equation}
%
and
%
\begin{align}
u(q')
&\equiv
e^{-\lambda \, q(t) + \lambda \, q'}.
\label{eq:ut_definition}
\end{align}
So
\begin{align}
&\left\langle x^2(t) \right\rangle
=
\langle x^2(0) \, \rangle e^{-2 \, \lambda \, q(t)}
\notag
\\
&\qquad
+
\int_0^t \int_0^t
  \dot u\bigl( q(t') \bigr) \,
  \dot u\bigl( q(t'') \bigr) \,
  \kappa(t' - t'') \, dt'' \, dt'.
\label{eq:x2t_average}
\end{align}

We now fix the endpoint value of $q(t)$
and variate the curve.
%
This is equivalent to variating $u$
with fixing endpoint values,
$u\bigl( q(0) \bigr)  = e^{-\lambda \, q(t)}$
and
$u\bigl( q(t) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:x2t_average} is fixed during
the variation process.
%
%We shall further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot u\bigl( q(t') \bigr) \,
%  \dot u\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
By the Euler-Lagrange equation, we get
$$
\begin{aligned}
0
&=
\frac{d}{d\tau} \int_0^t
  \dot u\bigl( q(t') \bigr) \, \kappa(t' - \tau) \, dt'
\\
&= \int_0^t
  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - \tau) \, dt',
\end{aligned}
$$
where we have dropped the boundary terms
by using Eq. \eqref{eq:kappat_limit}.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
In order for this to hold for any $\tau$,
we must have
%
\begin{equation}
\ddot u\bigl( q(t') \bigr) = 0,
\qquad
\mathrm{or}
\;\;
\dot u\bigl( q(t') \bigr) = c,
\label{eq:ddu_eq_0}
\end{equation}
%
where $c$ is a constant of $t'$.
%
Using Eq. \eqref{eq:ut_definition}
we get
$$
e^{-\lambda \, q(t) + \lambda \, q(t')}
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm, and differentiating this with respect to $t'$
yields Eq. \eqref{eq:alpha_opt}.

@@

Going back to the problem of
minimizing Eq. \eqref{eq:error_sum}
for the single-bin scheme.
%
Note that Eq. \eqref{eq:dxdt_WL}
is of the form of Eq. \eqref{eq:dxdt_alpha}.
%
Thus, with $\lambda = 1$,
Eq. \eqref{eq:alpha_opt}
will optimize each term $\langle x_i^2 \rangle$,
and hence the sum.
%
In other words, the $\alpha(t)$ given by
Eq. \eqref{eq:alpha_invt} is optimal
up to a constant $t_0$.

@@

The matrix $\mathbf w$ is not arbitrary.
%
A necessary condition to sample desired distribution
$\mathbf p = (p_1, \dots, p_n)$
is the following.
%
In Eq. \eqref{eq:vt_diffeq_mbin},
if $h_j(t)$ were the same as $p_j$,
the rates of change $\dot v_i(t)$'s
should be independent of $i$.
%
This allows $v_i(t)$'s to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1.
  \label{eq:w_sumj}
\end{equation}
%
The above condition means that $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
although the matrix elements can be negative.

To simplify the following discussion,
we shall further impose the
detailed balance condition:
%
\begin{equation}
  p_i \, w_{ij} = p_j \, w_{ji}.
  \label{eq:w_detailedbalance}
\end{equation}
%
It follows that
\begin{equation}
  \sum_{i = 1}^n p_i \, w_{ij}
  =
  \sum_{i = 1}^n p_j \, w_{ji}
  = p_j,
  \label{eq:w_balance}
\end{equation}
%
i.e., $\mathbf p$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.

Further, by Eq. \eqref{eq:w_detailedbalance},
we can define a symmetric matrix $\hat{\mathbf w}$
as $\hat w_{ij} = \sqrt{p_i/p_j} \, w_{ij}$
that can be diagonalized\cite{vankampen}
with a set of orthonormal eigenvectors:
%
$\sum_{i = 1}^n \varphi_{ki} \, \hat w_{ij} = \lambda_k \, \varphi_{kj}$.
%
Thus,
in terms of diagonalizing $\mathbf w$, we have
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij} = \lambda_k \, \phi_{kj},
  \label{eq:eig_w}
\end{equation}
where
$\phi_{ij} \equiv \sqrt{p_j} \, \varphi_{ij}$ satisfies
the orthonormal conditions\cite{vankampen}:
%
\begin{align}
\sum_{k = 1}^n \phi_{ki} \, \phi_{kj}
&= \delta_{ij} \, p_i,
\label{eq:eig_orthonormal_cols}
\\
\sum_{k = 1}^n \frac{ \phi_{ik} \, \phi_{jk} }{ p_k }
&= \delta_{ij}.
\label{eq:eig_orthonormal_rows}
\end{align}

%It is worth pointing out
Note that, in a stable updating scheme,
all eigenvalues $\lambda_k$ of $\mathbf w$
must be nonnegative,
for otherwise there is a fluctuating mode
that can increase indefinitely.

>>>>

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques
  in free energy calculations.
  %
  Both methods sample a flat distribution
  along a quantity of interest, $z$,
  by constructing a bias potential that offsets
  the potential of mean force (PMF) on the fly.
  %
  The two, however, differ by
  the manner of updating the bias potential:
  %
  in the WL case,
  each update applies to only
  the bin containing the current $z$,
  while in the metadynamics case,
  a Gaussian wave packet
  involving several neighboring bins
  is used.
  %
  The asymptotic error of the resulting PMF
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  the optimal schedule
  is known to be given by
  the inverse simulation time.
  %
  In this study, 
  we give a formula of computing the optimal schedule
  for a general updating scheme
  involving multiple bins.
  %
  We show that while metadynamics can be advantageous
  for relatively short simulations,
  the WL algorithm is best for asymptotic convergences.
  %the optimal schedule
  %and the effectiveness of metadynamics
  %are sensitive to the simulation length and other parameters.
  %
  %Further, we compared the two algorithms, and found that
  %although metadynamics produces a smoother PMF profile,
  %it is less effective in reducing errors with the shorter wavelengths.
  %%
  %Thus, the latter relies more heavily on the assumption of
  %the smoothness of the PMF profile,
  %especially with a wide updating window.
\end{abstract}

@@

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the problem is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$.
%
The negative logarithm,
$-\log p^*(z)$,
defines a free energy,
or the potential of mean force (PMF).
%
Usually, the distribution $p^*(z)$ is
localized in one or a few small $z$ regions.
%
Therefore, to capture the global shape
of the PMF,
one has to run multiple independent simulations
under different umbrella or bias potentials
along $z$\cite{mezei1987, berg1992, lee1993},
and then piece together the information.
%
This solution is clumsy in practice
and often suffers from a convergence problem
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum
for a long time.

A better solution is to artificially alter
the target distribution along $z$
to a flat one\cite{mezei1987, berg1992, lee1993,
wang2001, wang2001pre, laio2002, laio2008, barducci2011, sutto2012}.
%
This alleviates the above problem of local trapping
by forcing the system to travel more 
deliberately along $z$,
which often represents
a direction of slow motion.
%
To achieve a flat distribution, however,
the method must be able to construct a
bias potential that exactly cancels the PMF.

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and meta\-dy\-nam\-ics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
techniques for this purpose,
in which a bias potential is actively built-up on-the-fly
during simulation.
%
These techniques employ updating loops
that discourage the system
to sample configurations
sharing the current $z$ value
by elevating the bias potential here.
%
The main difference of the two
is that in the WL case,
each updating step is limited to the bin
containing the current $z$,
while metadynamics adopts an extended
updating wave packet
covering several neighboring bins.
%
The latter is more often used in MD simulations
as it guarantees a smooth profile
of the bias potential
suitable for numerical differentiation
in deriving the bias force.
%
However, in both cases, the bias potential
is updated frequently (usually every few MC or MD steps),
which brings about a side effect 
of disrupting the underlying
equilibrium sampling dynamics\cite{laio2002}.
%
Thus, one often has to reduce
the updating magnitude
over the course of simulation
to decrease the error.

The updating schedule,
or the manner of reducing
the updating magnitude\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014},
therefore,
determines the precision of the final bias potential,
hence that of the PMF.
%
For the WL algorithm, it is known
that the optimal magnitude of updating $\ln f$
is given by the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008},
where ``time'' means
the number of simulation steps
divided by the number of bins.
%
The optimal schedule for metadynamics,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general updating scheme,
including those used in WL and metadynamics simulations.
%
Although the method recovers the inverse-time formula
for the WL case,
it predicts, for metadynamics,
a more algebraically complex schedule,
which is sensitive to the simulation length
and the width of the Gaussian wave packet.
%
Further, by comparing
different updating schemes,
we show that
the single-bin updating scheme
used by the WL algorithm
is best for infinitely long simulations.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of
flat-distribution sampling
and the development on the optimal schedule.
%
This also helps fixing the notations.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
on the simplest case of
single-bin updating scheme.
%
We shall prove the optimality
of the known inverse-time formula.
%
Next, we turn to the general case
of multiple-bin updating schemes
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix}.
%
Finally, we compare different updating schemes
in the long time limit
in Sec. \ref{sec:cmpschemes}.

@@

\subsubsection{\label{sec:FDS}
Flat-distribution sampling}

@@

For a large system,
the distribution $p^*_i$ tends to
be localized,
%
and to find out the global shape,
it is often advantageous to carry out
a biased sampling that targets
a wider distribution $p_i$.
%
%Here, we refer to simulations that target
%a flat or nearly flat distribution
%as entropic or multicanonical sampling.

@@

Here we shall call a simulation that targets
a nearly flat distribution\cite{dayal2004, trebst2004, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS % (E-FDS) methods hereinafter,
estimates the bias potential, $V_i$,
and fixes it during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{berg1992, lee1993}
and simulated tempering\cite{marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, however,
can often be improved in the end of simulation.
%
To do so,
we accumulate a histogram along $i$
during simulation and use it for $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}.
%
This allows us to correct $V_i$ as
%
\begin{equation}
V_i^{\mathrm{corr}}
=
V_i^{\mathrm{}}
+
\log \frac{ \pi_i } { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
For an unknown system,
the main difficulty of the process
is that it is hard to guess a sufficiently accurate
$V_i$ to start with.
%
Thus, 
it usually takes several iterative rounds of simulations
to achieve the desired flat distribution.
%
Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.

For this reason, the second type of methods,
referred to as the adaptive FDS methods,
were invented.
%
Examples include
the WL and metadynamics methods.
%
These methods update the bias potential more frequently
and are more capable of achieving a flat distribution.
%
However, they are less well understood theoretically,
the updating loops used there
break the microscopic reversibility
of the underlying sampling mechanism.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods gradually
reduce the updating magnitude
over the course of simulation\cite{marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008}.
%
Thus, in late stages of a long simulation,
the updating is negligible and
such an adaptive simulation is rather similar
to the corresponding equilibrium simulation.
%
If the bias potential were converged to the exact value,
the equilibrium simulation can be thought as
the optimal asymptotic limit of the adaptive counterpart,
delivering the maximal possible efficiency.

@@

In the WL algorithm\cite{wang2001, wang2001pre},
$v_i$ are updated
in each MC step $t$,
and $v_i(t)$ as a function of $t$
is updated as
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }.
  \label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
We refer to the above scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme}, however,
several neighboring bins are updated:
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i },
  \label{eq:mbin_update}
\end{equation}
%
where $w_{ij}$ are a set of coefficients satisfying
$\sum_{j=1}^n w_{ij} = 1$.
%
We shall discuss other restrictions on $w_{ij}$
in Sec. \ref{sec:updating-matrix}.
%
The Gaussian updating scheme
used by metadynamics
is a multiple-bin scheme.

@@

In the original WL scheme,
the updating magnitude $\alpha$ (or $\ln f$
in the original paper) is kept as a constant
for a period of time,
which is referred to as a stage below.
%
The histogram collected in the stage is monitored.
%
Once the histogram is sufficiently flat,
we are allowed to enter a new stage
with a reduced $\alpha$\cite{wang2001, wang2001pre}
(usually half as large as
the previous value).
%
This scheme works well for early stages.
%
However, in later stages, it tends to reduce $\alpha$
too quickly, making the asymptotic error
saturate\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008}.
%
Similarly, for metadynamics,
there are various ways of reducing the updating magnitude,
particularly in the development of
self-healing umbrella sampling
and well-tempered metadynamics\cite{
  marsili2006, barducci2008, dickson2011, dama2014}.

@@

For the single-bin scheme
used in WL algorithm,
it was found that
a more effective way
of updating the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$ is the number of steps
from the beginning of the MC simulation,
which shall be referred to as the ``time'' below.
%
This surprisingly simple formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
There are, however, some reservation about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $C/t$ with a constant $C$
different from $1.0$\cite{
morozov2007, zhou2008, morozov2009}.

@@

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(T)
=
x(0) \, e^{-\lambda \, q(T)}
+
\int_0^T
  \dot u\bigl( q(t) \bigr) \, \xi(t) \, dt,
\label{eq:xt_solution}
\end{equation}
%
where,
%
\begin{equation}
q(t) \equiv \int_0^t \alpha(t') \, dt',
\label{eq:qt_definition}
\end{equation}
%
and
%
\begin{align}
u(q')
&\equiv
e^{-\lambda \, q(T) + \lambda \, q'}.
\label{eq:ut_definition}
\end{align}
At the end of period $T$, we have
\begin{align}
&\left\langle x^2(T) \right\rangle
=
\langle x^2(0) \,
\rangle e^{-2 \, \lambda \, q(T)}
\notag
\\
&\qquad
+
\int_0^T \int_0^T
  \dot u\bigl( q(t) \bigr) \,
  \dot u\bigl( q(t') \bigr) \,
  \kappa(t - t') \, dt' \, dt.
\label{eq:x2t_average}
\end{align}

We now fix the endpoint value of $q(T)$
and variate the curve.
%
This is equivalent to variating $u$
with fixing endpoint values,
$u\bigl( q(0) \bigr)  = e^{-\lambda \, q(T)}$
and
$u\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:x2t_average} is fixed during
the variation process.
%
%We shall further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot u(t') \, \dot u(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot u\bigl( q(t') \bigr) \,
%  \dot u\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
By the Euler-Lagrange equation, we get
$$
\begin{aligned}
0
&=
\frac{d}{dt} \int_0^T
  \dot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt'
\\
&= \int_0^T
  \ddot u\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt',
\end{aligned}
$$
where we have dropped the boundary terms
by using Eq. \eqref{eq:kappat_limit}.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t u^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
In order for this to hold for any $t$,
we must have
%
\begin{equation}
\ddot u\bigl( q(t') \bigr) = 0,
\qquad
\mathrm{or}
\;\;
\dot u\bigl( q(t') \bigr) = c,
\label{eq:ddu_eq_0}
\end{equation}
%
where $c$ is a constant of $t'$.
%
Using Eq. \eqref{eq:ut_definition},
we get
$$
e^{-\lambda \, q(t) + \lambda \, q(t')}
=
c \, (t' + t_0),
$$
where $t_0$ is a constant.
%
Taking the logarithm,
and differentiating both sides
with respect to $t'$
yields Eq. \eqref{eq:alpha_opt}.

@@

Going back to the problem of
minimizing Eq. \eqref{eq:error_sum}
for the single-bin scheme.
%
Note that Eq. \eqref{eq:dxdt_WL}
is of the form of Eq. \eqref{eq:dxdt_alpha}.
%
Thus, with $\lambda = 1$,
Eq. \eqref{eq:alpha_opt}
will optimize each term $\langle x_i^2 \rangle$,
and hence the sum.
%
In other words, the optimal $\alpha(t)$
is given by
\begin{equation}
\alpha(t)
=
\frac{    1     }
     {  t + t_0 }
,
\label{eq:alpha_invt1}
\end{equation}
agreeing with Eq. \eqref{eq:alpha_invt}.

@@

The matrix $\mathbf w$ is not arbitrary.
%
A necessary condition to sample
the desired distribution
$\mathbf p = (p_1, \dots, p_n)$
is the following.
%
In Eq. \eqref{eq:vt_diffeq_mbin},
if $h_j(t)$ were the same as $p_j$,
the rates of change $\dot v_i(t)$'s
should be independent of $i$.
%
This allows $v_i(t)$'s to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1.
  \label{eq:w_sumj}
\end{equation}
%
The above condition means that $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
although the matrix elements can be negative.

To simplify the ensuing discussion,
we shall further impose the
detailed balance condition:
%
\begin{equation}
  p_i \, w_{ij} = p_j \, w_{ji}.
  \label{eq:w_detailedbalance}
\end{equation}
%
It follows that
\begin{equation}
  \sum_{i = 1}^n p_i \, w_{ij}
  =
  \sum_{i = 1}^n p_j \, w_{ji}
  = p_j,
  \label{eq:w_balance}
\end{equation}
%
i.e., $\mathbf p$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.

Further, by Eq. \eqref{eq:w_detailedbalance},
we can define a symmetric matrix $\hat{\mathbf w}$
as $\hat w_{ij} = \sqrt{p_i/p_j} \, w_{ij}$
that can be diagonalized\cite{vankampen}
with a set of orthonormal eigenvectors:
%
$\sum_{i = 1}^n \varphi_{ki} \, \hat w_{ij} = \lambda_k \, \varphi_{kj}$.
%
Thus,
in terms of diagonalizing $\mathbf w$, we have
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij} = \lambda_k \, \phi_{kj},
  \label{eq:eig_w}
\end{equation}
where the eigenvectors
$$\phi_{ij} \equiv \sqrt{p_j} \, \varphi_{ij}$$ 
satisfy the orthonormal conditions\cite{vankampen}:
%
\begin{align}
\sum_{k = 1}^n \phi_{ki} \, \phi_{kj}
&= \delta_{ij} \, p_i,
\label{eq:eig_orthonormal_cols}
\\
\sum_{k = 1}^n \frac{ \phi_{ik} \, \phi_{jk} }{ p_k }
&= \delta_{ij}.
\label{eq:eig_orthonormal_rows}
\end{align}

Particularly, there is always an eigenvector 
corresponding to the equilibrium distribution.
%
We shall label this eigenvector as the first one,
and
%
\begin{equation}
\phi_{1k} = p_k,
\label{eq:eigenmode1}
\end{equation}
%
which satisfies Eq. \eqref{eq:eig_orthonormal_rows}.
%
The corresponding eigenvalue is $\lambda_1 = 1$.

%It is worth pointing out
Further, in a stable updating scheme,
all eigenvalues $\lambda_k$ of $\mathbf w$
must be nonnegative,
for otherwise there is a fluctuating mode
that can increase indefinitely.
